---
title: "Fear of Flying - Risk Evaluation"
format: html
df-print: kable
editor: visual
output-file: index.html
---

## Fear of Flying

Many people are fearful of flying on commercial passenger airplanes for a multitude of reasons. More specifically, people may experience fear regarding whether their flight will have safe passage. Why are people fearful? After all, if you conduct a quick internet search, it appears that flying is considered the safest method of transportation.

Let's take a deeper dive into what the data says.

## What Does the Data Say?

To analyze how safe commercial passenger airplanes are, one approach is to get the total number of commercial passenger flights that have occurred over a given period of time. The region in which this analysis will occur is the United States, where both the departure and arrival locations are in the United States or where either the departure or arrival locations are in the United States.

Moreover, we can narrow down the fear of flying to be more specific. If you are located in the United States, you are likely to fly on the most popular airline carriers, such as:

-   JetBlue Airways

-   Southwest Airlines

-   American Airlines

-   Delta Airlines

-   Continental Airlines

-   Frontier Airlines

-   SkyWest Airlines

-   Allegiant Airlines

-   Spirit Airlines

-   United Airlines

So, this analysis will focus on the ten (10) carriers mentioned above.

### Total Flights by Air Carrier

Let's take a look at the total number of flights by air carrier. The date range selected was from January 2008 - August 2024. Flight totals are presented by month. The total number of flights for each airline was collected from the Bureau of Transportation Statistics (BTS). Below is a table that shows the total number of flights by airline within this time period:

```{r}
#| echo: false
#| warning: false

library(tidyverse)

load("airtraffic.Rdata")
load("aircraft_events.Rdata")

# summary stats on airtraffic data --------------------------------------------

airtraffic_sum_stats <- airtraffic %>%
  group_by(unique_carrier_name, year) %>%
  summarize(
    num_flights = n()
  ) %>%
  rename(airline_name = unique_carrier_name) %>%
  ungroup()

# summary stats on aircraft_events data ---------------------------------------

aircraft_events_sum_stats <- aircraft_events %>%
  # filter is applied to match the same time period as airtraffic
  filter(!(year == 2024 & month > 8)) %>%
  group_by(owner_acft, year, month) %>% summarize(
    num_aircraft_in_incident = n()
  ) %>% 
  rename(airline_name = owner_acft) %>%
  ungroup()

## create temp df to join to sum stat -----------------------------------------

# the temp df includes all months of all years of interest for all airlines,
# even if there are no records for that month
timeframe <- seq(from = as.Date("2008-01-01"), to = as.Date("2024-12-01"), by = "month")
airline_name <- unique(aircraft_events$owner_acft)

temp_df <- data.frame(
  
  airline_name = rep(airline_name, each = length(timeframe)),
  timeframe = rep(timeframe, times = length(airline_name))

) %>% 
  mutate(
  year = year(timeframe),
  month = month(timeframe)) %>%
  select(-timeframe)

aircraft_events_sum_stats <- temp_df %>%
  left_join(., aircraft_events_sum_stats, by = join_by(airline_name, year, month)) %>%
  mutate(num_aircraft_in_incident = if_else(is.na(num_aircraft_in_incident), 0, num_aircraft_in_incident))

aircraft_events_sum_stats <- aircraft_events_sum_stats %>%
  filter(!(year == 2024 & month > 8))

# combined stats -------------------------------------------------------------

combined_stat_month <- airtraffic_sum_stats %>%
  left_join(., aircraft_events_sum_stats) %>%
  mutate(num_aircraft_in_incident = case_when(
    
    is.na(num_aircraft_in_incident) ~ 0,
    TRUE ~ num_aircraft_in_incident
    
  ))

combined_stat_year <- combined_stat_month %>%
  group_by(airline_name, year) %>% 
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  )

combined_total <- combined_stat_month %>% 
  group_by(airline_name) %>%
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  ) %>%
  mutate(prcnt_incid = (total_incidents/total_flights) * 100)

```

```{r}
#| echo: false
#| warning: false

total_flights <- combined_total %>%
  select(-c(prcnt_incid, total_incidents))

total_flights

```

\

To further illustrate the total number of flights by airline over time from January 2008 - August 2024, a graph is included below:

```{r}
#| echo: false
#| warning: false

p <- ggplot(airtraffic_sum_stats, aes(year, num_flights, group = airline_name, color = airline_name)) +
  geom_line(linewidth = 1) +
  theme_bw() + 
  labs(
    y = "Number of Flights",
    color = "Airline Name"
  ) +
  scale_x_continuous(limits = c(2008, 2025)) +
  scale_y_continuous(limits = c(0, 60000)) +
  theme(axis.title.x = element_blank()) + 
  theme(axis.title.y = element_text(margin = margin(r=10)))


p


```

### Total Incidents by Air Carrier

Now, let's take a look at the number of "events" reported by the National Transportation Safety Board (NTSB). The NTSB, from an aviation perspective, is tasked with investigating civil cases of airplane accidents/incidents. The accidents/incidents vary in severity and are dependent on the level of damage sustained by the aircraft and the level of injury sustained by passengers. Here are the number of incidents investigated by the NTSB from the timeframe of January 2008 - August 2024:

```{r}
#| echo: false
#| warning: false

aircraft_events_sum_stats_all <- aircraft_events_sum_stats %>%
  group_by(airline_name) %>%
  summarize(num_incidents = sum(num_aircraft_in_incident))

aircraft_events_sum_stats_all


```

Here are the number of incidents that have occurred over time, organized by airline:

```{r}
#| echo: false
#| warning: false

aircraft_events_sum_stats_year <- aircraft_events_sum_stats %>%
  group_by(airline_name, year) %>%
  summarize(num_incidents = sum(num_aircraft_in_incident))


p <- ggplot(aircraft_events_sum_stats_year, aes(year, num_incidents)) +
  geom_line() +
  theme_bw() + 
  labs(
    y = "Number of Aircraft in Incident",
    color = "Airline Name"
  ) +
  scale_x_continuous(limits = c(2008, 2025)) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(axis.title.x = element_blank()) + 
  theme(axis.title.y = element_text(margin = margin(r=10)))

p <- p + facet_wrap(vars(airline_name))

p


```

As the data shows above, some airlines have experienced more incidents over time. Let's normalize the number of incidents endured by each airline by the number of each airline's flights. The last column in the table below - the "prcnt_incid" column - represents each airline's rate of their respective craft experiencing an incident.

```{r}
#| echo: false
#| warning: false


combined_total


```

Now, let's look at how the "prcnt_incid" field is displayed in a graphical format:

```{r}
#| echo: false
#| warning: false

p <- ggplot(combined_total, aes(airline_name, prcnt_incid, fill = airline_name)) +
  geom_bar(stat = "identity", show.legend = FALSE) + 
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  labs(y = "Percent Incident Rate") +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(axis.text.x = element_text(vjust = 0.5))
  

p



```

Here are summary statistics across all airline's ratio of aircraft involved in an incident relative to the total flights flown:

```{r}
#| echo: false
#| warning: false

sum_stats <- combined_total %>%
  summarize(
    min = min(prcnt_incid),
    max = max(prcnt_incid),
    mean = mean(prcnt_incid),
    median = median(prcnt_incid)
  )

sum_stats



```

## Types of Incidents

It is important to remember that the NTSB data shown above is inclusive of all types of incidents. Some of these incidents cause no harm to the passengers on board, while other incidents can be more serious. ***For example, among the airlines involved in this study, there are only 2 incidents out of 3,084,159 flights that resulted in fatalities between January 2008 - August 2024, which corresponds to a 0.000065 % rate that these U.S. air carriers collectively experience a fatal incident.***

\*\*\*This means that, out of the 3,084,159 flights involved in this study, 2

```{r}
#| echo: false
#| warning: false

all_events_sumstats <- aircraft_events %>%
  group_by(ev_highest_injury) %>%
  summarize(
    
    Count = n()
    
  ) %>%
  rename(`Highest Injury` = ev_highest_injury) %>%
  mutate(`Highest Injury` = case_when(
    
    `Highest Injury` == "FATL" ~ "Fatal",
    `Highest Injury` == "MINR" ~ "Minor",
    `Highest Injury` == "SERS" ~ "Serious",
    `Highest Injury` == "NONE" ~ "None",
    is.na(`Highest Injury`) ~ "NA",
    TRUE ~ NA
    
    
  ))

all_events_sumstats

```

## Conclusions

Some conclusions that come from the data above are the following:

1.  The airline with the highest ratio (0.002 %) of their aircraft involved in incidents relative to the number of flights flown (hereinafter referred to as the "ratio") is American Airlines.
2.  The average ratio of all airlines part of this study is 0.00084 %.
3.  Even with this ratio, this means there is a 1 in 50,000 chance that a plane will experience an *incident*; again, an incident can have a variety of different meanings that can range from minor aircraft damage to crashes.

## Notes on the Data

-   The NTSB data covers incidents that occur by US air carriers in both domestic and international locations. The BTS data includes both domestic and international locations as well; however, if the flight's origin and destination are both located on international ground, then this is excluded. See the "T-100 Segment (US Carriers Only)" table located here: [https://www.transtats.bts.gov/Tables.asp?QO_VQ=EED&QO_anzr=Nv4%FDPn44vr4%FDf6n6v56vp5%FD%FLS14z%FDHE%FDg4nssvp%FM-%FD%FDh.f.%FDPn44vr45&QO_fu146_anzr=Nv4%FDPn44vr45](https://www.transtats.bts.gov/Tables.asp?QO_VQ=EED&QO_anzr=Nv4%FDPn44vr4%FDf6n6v56vp5%FD%FLS14z%FDHE%FDg4nssvp%FM-%FD%FDh.f.%FDPn44vr45&QO_fu146_anzr=Nv4%FDPn44vr45).

-   NTSB data can be obtained here: <https://data.ntsb.gov/avdata>
