---
title: "Fear of Flying - Risk Evaluation"
format: html
editor: visual
output-file: test.html
---

## Fear of Flying

Many people are fearful of flying on commercial passenger airplanes for a multitude of reasons. More specifically, people may experience fear regarding whether their flight will have safe passage. Why are people fearful? After all, if you conduct a quick internet search, it appears that flying is considered the safest method of transportation.

Let's take a deeper dive into what the data says.

## What Does the Data Say?

To analyze how safe commercial passenger airplanes are, one approach is to get the total number of commercial passenger flights that have occurred over a given period of time. The region in which this analysis will occur is the United States, where both the departure and arrival locations are in the United States or where either the departure or arrival locations are in the United States.

Moreover, we can narrow down the fear of flying to be more specific. If you are located in the United States, you are likely to fly on the most popular airline carriers, such as:

-   JetBlue Airways

-   Southwest Airlines

-   American Airlines

-   Delta Airlines

-   Continental Airlines

-   Frontier Airlines

-   SkyWest Airlines

-   Allegiant Airlines

-   Spirit Airlines

-   United Airlines

So, this analysis will focus on the ten (10) carriers mentioned above.

### Total Flights by Air Carrier

Let's take a look at the total number of flights by air carrier. The date range selected was from January 2008 - November 2024. Flight totals are presented by month.

```{r}
#| echo: false
#| warning: false

library(tidyverse)

load("airtraffic.Rdata")
load("aircraft_events.Rdata")

# summary stats on airtraffic data --------------------------------------------

airtraffic_sum_stats <- airtraffic %>%
  group_by(unique_carrier_name, year) %>%
  summarize(
    num_flights = n()
  ) %>%
  rename(airline_name = unique_carrier_name) %>%
  ungroup()

p <- ggplot(airtraffic_sum_stats, aes(year, num_flights, group = airline_name, color = airline_name)) +
  geom_line(linewidth = 1) +
  theme_bw() + 
  labs(
    y = "Number of Flights",
    color = "Airline Name"
  ) +
  theme(axis.title.x = element_blank()) + 
  theme(axis.title.y = element_text(margin = margin(r=10)))

p

# summary stats on aircraft_events data ---------------------------------------

aircraft_events_sum_stats <- aircraft_events %>%
  group_by(owner_acft, year, month) %>% summarize(
    num_aircraft_in_incident = n()
  ) %>% 
  rename(airline_name = owner_acft) %>%
  ungroup()

## create temp df to join to sum stat -----------------------------------------

# the temp df includes all months of all years of interest for all airlines,
# even if there are no records for that month
timeframe <- seq(from = as.Date("2008-01-01"), to = as.Date("2024-12-01"), by = "month")
airline_name <- unique(aircraft_events$owner_acft)

temp_df <- data.frame(
  
  airline_name = rep(airline_name, each = length(timeframe)),
  timeframe = rep(timeframe, times = length(airline_name))

) %>% 
  mutate(
  year = year(timeframe),
  month = month(timeframe)) %>%
  select(-timeframe)

aircraft_events_sum_stats <- temp_df %>%
  left_join(., aircraft_events_sum_stats, by = join_by(airline_name, year, month)) %>%
  mutate(num_aircraft_in_incident = if_else(is.na(num_aircraft_in_incident), 0, num_aircraft_in_incident))

# combined stats -------------------------------------------------------------

combined_stat_month <- airtraffic_sum_stats %>%
  left_join(., aircraft_events_sum_stats) %>%
  mutate(num_aircraft_in_incident = case_when(
    
    is.na(num_aircraft_in_incident) ~ 0,
    TRUE ~ num_aircraft_in_incident
    
  ))

combined_stat_year <- combined_stat_month %>%
  group_by(airline_name, year) %>% 
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  )

combined_total <- combined_stat_month %>% 
  group_by(airline_name) %>%
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  ) %>%
  mutate(prcnt_incid = (total_incidents/total_flights) * 100)

```

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
