---
title: "How Safe is Flying?"
format: html
#df-print: flextable
editor: visual
output-file: index.html
---

## The Question

You may be asking "How Safe is Flying?" (hereinafter referred to as the "Question") for a multitude of reasons, some of which may be that you are fearful of flying or you are genuinely curious how many airplane accidents have occurred over a given period of time. Regardless, the research presented herein attempts to answer this Question.

### Assumptions:

Firstly, let's try to be more specific about the question we are trying to answer. We will assume that when we ask the Question, the Question assumes we are referring to flights that involve popular U.S. commercial airlines. For this study, we will consider popular U.S. commercial airlines as the following:

-   JetBlue Airways

-   Southwest Airlines

-   American Airlines

-   Delta Airlines

-   Continental Airlines

-   Frontier Airlines

-   SkyWest Airlines

-   Allegiant Airlines

-   Spirit Airlines

-   United Airlines

## The Approach: How to Answer the Question

One intuitive method to answer the Question is to determine the number of airplane accidents that have occurred for each of the listed U.S. commercial airlines relative to the number of flights each of the listed U.S. commercial airlines have had over a given period of time. This approach will allow us to determine a rate in which airplane accidents occur for each airline. The data could also be combined to provide a collective rate in which airplane accidents occur across all airlines.

## What Does the Data Say?

### Total Flights

Based on our approach, let's first take a look at the total number of flights by airline.

The Bureau of Transportation Statistics (BTS) provides air traffic data, including flight data for U.S. commercial airlines. Let's use this data as a method of determining each airline's number of flights over a given period of time. For the purpose of this study, we will focus on the time range of January 2008 - August 2024, hereinafter referred to as the "time period".

The data compiled from the BTS can be obtained [here](https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=GDM&QO_fu146_anzr=Nv4%20Pn44vr45). Note that the flight data includes all flights that fly domestically within the United States and also includes international flights where either the destination or departure location are located in the United States.

Below is a table that shows the total number of flights by airline within the provided time period:

```{r}
#| echo: false
#| warning: false

library(tidyverse)
library(knitr)
library(flextable)


# function for formatting flextables ------------------------------------------

df_formatting <- function(df) {
  
  ft <- flextable(df)
  
  ft <- ft %>% 
    theme_vanilla() %>%
    vline() %>%
    border_outer() %>%
    align(align = "center", part = "all") %>%
    flextable::bg(bg = "grey", part = "header") %>%
    autofit()

  return(ft)  
  
  
}


# load data -------------------------------------------------------------------

load("airtraffic.Rdata")
load("aircraft_events.Rdata")

# summary stats on airtraffic data --------------------------------------------

airtraffic_sum_stats <- airtraffic %>%
  group_by(unique_carrier_name, year) %>%
  summarize(
    num_flights = n()
  ) %>%
  rename(airline_name = unique_carrier_name) %>%
  ungroup()

# summary stats on aircraft_events data ---------------------------------------

aircraft_events_sum_stats <- aircraft_events %>%
  # filter is applied to match the same time period as airtraffic
  filter(!(year == 2024 & month > 8)) %>%
  group_by(owner_acft, year, month) %>% summarize(
    num_aircraft_in_incident = n()
  ) %>% 
  rename(airline_name = owner_acft) %>%
  ungroup()

## create temp df to join to sum stat -----------------------------------------

# the temp df includes all months of all years of interest for all airlines,
# even if there are no records for that month
timeframe <- seq(from = as.Date("2008-01-01"), to = as.Date("2024-12-01"), by = "month")
airline_name <- c(unique(aircraft_events$owner_acft), "allegiant airlines")

temp_df <- data.frame(
  
  airline_name = rep(airline_name, each = length(timeframe)),
  timeframe = rep(timeframe, times = length(airline_name))

) %>% 
  mutate(
  year = year(timeframe),
  month = month(timeframe)) %>%
  select(-timeframe)

aircraft_events_sum_stats <- temp_df %>%
  left_join(., aircraft_events_sum_stats, by = join_by(airline_name, year, month)) %>%
  mutate(num_aircraft_in_incident = if_else(is.na(num_aircraft_in_incident), 0, num_aircraft_in_incident))

aircraft_events_sum_stats <- aircraft_events_sum_stats %>%
  filter(!(year == 2024 & month > 8))

# combined stats -------------------------------------------------------------

combined_stat_month <- airtraffic_sum_stats %>%
  left_join(., aircraft_events_sum_stats) %>%
  mutate(num_aircraft_in_incident = case_when(
    
    is.na(num_aircraft_in_incident) ~ 0,
    TRUE ~ num_aircraft_in_incident
    
  ))

combined_stat_year <- combined_stat_month %>%
  group_by(airline_name, year) %>% 
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  )

combined_total <- combined_stat_month %>% 
  group_by(airline_name) %>%
  summarize(
    total_flights = sum(num_flights),
    total_incidents = sum(num_aircraft_in_incident)
  ) %>%
  mutate(prcnt_incid = (total_incidents/total_flights) * 100)

```

```{r}
#| echo: false
#| warning: false


total_flights <- combined_total %>%
  select(-c(prcnt_incid, total_incidents)) %>%
  rename(`Airline Name` = airline_name,
         `Total Flights` = total_flights)

df_formatting(total_flights)

```

**Based on the above table, the total number of flights across all airlines equals 36,224,208.**

It may be useful to visualize the number of flights by airline in graphical format to see which airlines are more popular than others. As such, a graph displaying this information is provided below:

```{r}
#| echo: false
#| warning: false


p <- ggplot(airtraffic_sum_stats, aes(year, num_flights, group = airline_name, color = airline_name)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    y = "Number of Flights",
    color = "Airline Name"
  ) +
  scale_x_continuous(limits = c(2008, 2025)) +
  scale_y_continuous(limits = c(0, 60000)) +
  theme(axis.title.x = element_blank()) +
  theme(axis.title.y = element_text(margin = margin(r=10)))

p


```

::: callout-note
Note that, for the graphic above, not all flights through the end of 2024 were collected since the time range for this study ends at August 2024.
:::

### Total Events

Now let's take a look at the number of airline accidents that have occurred over the same time period.

The data that will be used to get this information is derived from the National Transportation Safety Board (NTSB); the NTSB, from an aviation perspective, is tasked with investigating civil cases of airplane accidents/incidents, otherwise referred to as "events". The events vary in severity and are dependent on the level of damage sustained by the aircraft and the level of injury sustained by passengers. The data can be downloaded [here](https://data.ntsb.gov/avdata).

Below are the number of incidents investigated by the NTSB relative to this study's time period:

```{r}
#| echo: false
#| warning: false

aircraft_events_sum_stats_all <- aircraft_events_sum_stats %>%
  group_by(airline_name) %>%
  summarize(`Number of Events` = sum(num_aircraft_in_incident)) %>%
  rename(`Airline Name` = airline_name)

df_formatting(aircraft_events_sum_stats_all)


```

It may be useful to visualize the number of incidents by airline in a graphical format. As such, a graph displaying this information is provided below:

```{r}
#| echo: false
#| warning: false

aircraft_events_sum_stats_year <- aircraft_events_sum_stats %>%
  group_by(airline_name, year) %>%
  summarize(num_incidents = sum(num_aircraft_in_incident))


p <- ggplot(aircraft_events_sum_stats_year, aes(year, num_incidents)) +
  geom_line() +
  theme_bw() + 
  labs(
    y = "Number of Aircraft in an NTSB Event",
    color = "Airline Name"
  ) +
  scale_x_continuous(limits = c(2008, 2025)) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(axis.title.x = element_blank()) + 
  theme(axis.title.y = element_text(margin = margin(r=10)))

p <- p + facet_wrap(vars(airline_name))

p


```

### Events vs. Flights

Now that we have the total number of flights and total number of events, we can determine the rate in which an event occurs for each airline.

Below is a table that shows the total number of flights, the total number of events, and the percent rate in which an event occurs for each airline.

```{r}
#| echo: false
#| warning: false


combined_total_display <- combined_total %>%
  rename(`Airline Name` = airline_name,
         `Total Flights` = total_flights,
         `Total Events` = total_incidents,
         `% Event Rate` = prcnt_incid)

df_formatting(combined_total_display)


```

Here are summary statistics on the data contained within the "% Event Rate" column shown above:

```{r}
#| echo: false
#| warning: false

sum_stats <- combined_total %>%
  summarize(
    min = min(prcnt_incid),
    max = max(prcnt_incid),
    mean = mean(prcnt_incid),
    median = median(prcnt_incid)
  )

df_formatting(sum_stats)



```

Now, let's look at how the "% Event Rate" field is displayed in a graphical format for easier comparison:

```{r}
#| echo: false
#| warning: false

p <- ggplot(combined_total, aes(airline_name, prcnt_incid, fill = airline_name)) +
  geom_bar(stat = "identity", show.legend = FALSE) + 
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  labs(y = "Percent NTSB Event Rate") +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(axis.text.x = element_text(vjust = 0.5))
  

p



```

### Types of Events

It is important to note that the NTSB data shown above is inclusive of all types of events. Some of these events cause no harm to the passengers on board, while other events can be more serious. The NTSB categorizes the level of injury incurred by an event as fatal, serious, minor, none, or NA. The table below shows how the events across the time period of this study are categorized:

```{r}
#| echo: false
#| warning: false

all_events_sumstats <- aircraft_events %>%
  group_by(ev_highest_injury) %>%
  summarize(
    
    Count = n()
    
  ) %>%
  rename(`Highest Injury` = ev_highest_injury) %>%
  mutate(`Highest Injury` = case_when(
    
    `Highest Injury` == "FATL" ~ "Fatal",
    `Highest Injury` == "MINR" ~ "Minor",
    `Highest Injury` == "SERS" ~ "Serious",
    `Highest Injury` == "NONE" ~ "None",
    is.na(`Highest Injury`) ~ "NA",
    TRUE ~ NA
    
    
  ))

df_formatting(all_events_sumstats)

```

Based on the table above, ***among the airlines involved in this study, there are only 2 events out of the 36,224,208 flights that resulted in fatalities between January 2008 - August 2024, which corresponds to a 0.00000552 % rate that these U.S. airlines collectively experience a fatal incident.***

What about the NA values in the "Highest Injury" column, though? To ensure there aren't unaccounted for flights that endured some level of passenger injury, we can take a look at the incidents that have an NA value in the "Highest Injury" column by investigating the NTSB narratives for each event. The narratives are lengthy, but some examples of these narratives are included below:

Example 1:

```{r}
#| echo: false
#| warning: false

na_narratives <- aircraft_events %>%
  filter(is.na(ev_highest_injury)) %>%
  select(ev_id, narr_accp) %>%
  rename(`Event ID` = ev_id,
         Narrative = narr_accp)


x <- as.vector(na_narratives[1, 2])

x

```

Example 2:

```{r}
#| echo: false
#| warning: false

na_narratives <- aircraft_events %>%
  filter(is.na(ev_highest_injury)) %>%
  select(ev_id, narr_accp) %>%
  rename(`Event ID` = ev_id,
         Narrative = narr_accp)


x <- as.vector(na_narratives[13, 2])

x

```

Upon review of the narratives where the "Highest Injury" value is NA, there did not appear to be an explicit indication that any injuries occurred; however, this is ultimately inconclusive for this study.

## Conclusions

Some conclusions that come from the data above are the following:

1.  The airline with the highest ratio (0.002 %) of their aircraft involved in an NTSB investigated event relative to the number of flights flown is American Airlines. Even with this ratio, this means there is a 1 in 50,000 chance that an American Airlines plane will experience an *incident*; again, an incident can have a variety of different meanings that can range from minor aircraft damage to crashes.

2.  The average ratio of aircraft involved in an NTSB event relative to the total flights flown for each U.S. airline part of this study is 0.00084 %.

3.  Among the U.S. airlines involved in this study, there were only 2 flights out of 36,224,208 that resulted in fatalities between January 2008 - August 2024; this corresponds to a 0.00000552 % rate (1 in every 18,112,104 flights) that the U.S. airlines involved in this study collectively experience a fatal incident. 

4. Among the U.S. airlines involved in this study, there were 151 flights out of 36,224,208 that resulted in fatalities or a serious injury between January 2008 - August 2024; this corresponds to a 0.000417 % rate (1 in every 239,895 flights) that the U.S. airlines involved in this study collectively experience an incident resulting in fatal or serious injuries. 

## Answering the Question

The conclusions formulated above can be used to answer the Question. Based on the data discussed herein, flying on U.S. commercial airlines - at least those that are part of this study - is safe since the data has shown a relatively low probability that any given flight will be involved in an NTSB event and a much smaller probability that any given flight will be involved in an NTSB event resulting in fatalities. 

## Notes on the Data

A few assumptions are made about the data used in this research. One assumption is that we assume the BTS data accurately records every domestic and international flight for all the airlines referenced herein. Again, international in this context means a flight where the destination or departing location is in the U.S., while the other location is in a foreign county. Secondly, we assume that the NTSB records the correct owner of the aircraft involved in an event for each of the airlines referenced in this study. 
