---
title: "Fear of Flying - Risk Evaluation"
format: html
editor: visual
---
# test
\

## Fear of Flying

(Insert language that expounds on the fear of flying).

## What Does the Data Say?

```{r}
library(tidyverse)
library(readxl)
library(janitor)


# data taken from here: https://www.kaggle.com/datasets/yyxian/u-s-airline-traffic-data/code
airtraffic <- read_csv("C:\\Users\\joefo\\Files\\coding_projects\\flying_risk_eval\\data\\airtraffic.csv")

airtraffic2 <- airtraffic %>% 
  janitor::clean_names() %>%
  select(year, month, pax, flt) %>%
  rename(passengers_total = pax, flights_total = flt)
  
crashdata <- read_csv("C:\\Users\\joefo\\Files\\coding_projects\\flying_risk_eval\\data\\crashdata.csv")

crashdata2 <- crashdata %>%
  janitor::clean_names()

crashdata2 <- crashdata2 %>%
  mutate(event_date = date(event_date)) %>%
  filter(year(event_date) >= 2003) %>%
  select(event_date, city, state, country, highest_injury_level:probable_cause, make, model)

unique_values <- lapply(colnames(crashdata2), function(elem){
  
 x <- unique(crashdata2[[elem]])
 return(x)
  
})

names(unique_values) <- colnames(crashdata2)

test <- crashdata2 %>% 
  group_by(make, model) %>% 
  filter(str_detect(tolower(make), "boeing|airbus")) %>% 
  summarise(n = n())

```

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

Thi
